A cascaded supervised learning approach to inverse reinforcement learning

### 2012-11-07 piot2012cascading/ECML13_cascading.tex
Dernieres modifs avant soumission

% Created 2012-10-15 lun. 14:49
\documentclass[smallextended]{svjour3}
\smartqed 
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage[integrals]{wasysym}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subfigure}
\tolerance=1000
%\usepackage{amsmath}
%\usepackage{amsthm}
%\newtheorem{definition}{Definition}
%\newtheorem{theorem}{Theorem}
\newtheorem{cor}{Corollary}
\usepackage{dsfont}
\usepackage{mathtools}
\mathtoolsset{showonlyrefs=true}
\providecommand{\alert}[1]{\textbf{#1}}
\newcommand{\E}{\mathbf{E}}
\newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
\newcommand{\Card}{\operatorname*{Card}} %\operatorname* pour les op. pouvant admettre des limites...
\begin{document}
\title{A cascaded supervised learning approach to inverse reinforcement learning}
\author{Edouard Klein$^{1,2}$ \and Bilal Piot $^{2,3}$\and Matthieu Geist $^{2}$\and Olivier Pietquin$^{2,3}$}
\titlerunning{Cascading approach to IRL}
\institute{
 1. ABC Team\\
 LORIA-CNRS, France.\\
\and
2.Supélec\\
 IMS-MaLIS Research group, France\\
 \texttt{firstname.lastname@supelec.fr}\\
\and 
3. UMI 2958 GeorgiaTech-CNRS\\
France\\
}
\date{\today}

@@

\begin{abstract}
  This paper considers the Inverse Reinforcement Learning (IRL) problem, that is inferring a reward function for which a demonstrated expert policy is optimal.
We propose to break the IRL problem down into two generic Supervised Learning steps: this is the Cascaded Supervised IRL (CSI) approach. A classification step that defines a score function is followed by a regression step providing a reward function.
A theoretical analysis shows that the demonstrated expert policy is near-optimal for the computed reward function.
Not needing to repeatedly solve a Markov Decision Process (MDP) and the ability to leverage existing techniques for classification and regression are two important advantages of the CSI approach. It is furthermore empirically demonstrated to work with only transitions sampled according to the expert policy, up to the use of some heuristics. This is exemplified on two classical benchmarks (a highway driving simulator and an inverted pendulum simulator).
  \end{abstract}
\section{Introduction}
\label{sec-2}
Sequential decision making consists in choosing the appropriate action given the available data in order to maximize a certain criterion. When framed in a Markov Decision Process (MDP) (see Sec.~\ref{sec:background}), the criterion to maximize is the expected sum of discounted rewards. The Inverse Reinforcement Learning (IRL)~\cite{russell1998learning} problem, which we tackle here, aims at inferring a reward function for which a demonstrated expert policy is optimal.

IRL is often used as a way to do Apprenticeshipe Learning (AL): imitating a demonstrated expert policy, without explicitely looking for the reward function. The reward function nevertheless is of interest in and of itself. As mentionned in \cite{ng2000algorithms}, its semantics can be analyzed in biology or econometrics. Practically, the reward can be seen as a succinct description of a task. Discovering it removes the coupling between understanding the task and learning how to fulfill it that exists in AL. IRL allows the use of (Approximate) Dynamic programming ((A)DP) or Reinforcement Learning (RL) techniques to learn the task from the computed reward function. Although some approaches use it as a byproduct, most AL algorithm make not use of the reward function. A very straightforward way to do AL is to use a multi-class classifier to directly learn the expert policy. We compare this idea with our IRL algorithm in the experiments (Sec.~\ref{sec:experiments}).

A lot of existing approaches in either IRL or AL need to repeatedly solve the underlying MDP. Thus, their performance depends strongly on the quality of the associated subroutine. Consequently, they suffer from the same challenges of scalability, data scarcity, etc., as RL and (A)DP. In order to avoid repeatedly solving such problems, we adopt a different point of view.

Having in mind that there is a one to one relation between a reward function and its associated optimal action-value function (via the Bellman equation, see Eq.~\eqref{eq:bellman1}), it is worth thinking of a method able to output an action-value function for which the greedy policy is the demonstrated expert policy. Thus, the demonstrated expert policy will be optimal for the related reward function. We propose a score function-based multi-class classification step (see Sec.~\ref{sec:algo}) to infer a score function. Besides, in order to retrieve via the Bellman equation the reward associated to the score function computed by the classification step, we introduce a regression step (see Sec.~\ref{sec:algo}). That is why the method is called the Cascaded Supervised Inverse reinforcement learning (CSI). This method is analyzed, see Sec.~\ref{sec:analysis} where it is shown that the demonstrated expert policy is near-optimal for the reward the regression step outputs.

This algorithm does not need to iteratively solve an MDP and needs only sampled transitions from expert and non-experts policies as inputs. Moreover, up to the use of some heuristics (see Sec.~\ref{subsubsec:heuristics}), the algorithm is able to be trained only with transitions sampled from the demonstrated expert policy. A specific instantiation of CSI (proposed Sec.~\ref{subsec:instanciation}) is tested on an inverted pendulum simulator (Sec.~\ref{subsec:IP}) and a car driving simulator (Sec.~\ref{subsec:Highway}) where we compare it with an AL algorithm \cite{abbeel2004apprenticeship}, with a pure classification algorithm \cite{taskar2005learning} and with a recent successfull IRL method \cite{klein2012scirl} as well as with a random baseline.

@@

We work with finite MDPs \cite{puterman1994markov}, that is tuples $\{S,A,P,R,\gamma\}$. The state space is noted $S$, $A$ is a finite action space, $R\in\mathbb{R}^{S\times A}$ is a reward-function, $\gamma\in ]0,1[$ is a discount factor and $P\in \Delta_{S}^{S\times A}$ is the Markovian dynamic of the MDP. Thus for each $(s,a)\in S\times A$, $P(.|s,a)$ is a distribution over $S$ and $P(s'|s,a)$ is the probability to reach $s'$ by choosing the action $a$ in the state $s$. At each time step $t$, the agent uses the information encoded in the state $s_t\in S$ in order to choose an action $a_t \in A$ according to a (deterministic\footnote{We restrict ourselves here to deterministic policies, but the loss of generality is minimal as there exists at least one optimal deterministic policy.}) policy $\pi\in A^S$. The agent then steps to a new state $s_{t+1}\in S$ according to the Markovian transition probabilities $P(s_{t+1}|s_t,a_t)$. The stationary distribution over the states $\rho_\pi$ induced by a policy $\pi$ satisfies
\begin{equation}
  \rho_\pi^TP_\pi = \rho_\pi^T,
  \end{equation}
where $P_\pi$ is $(P(s'|s,\pi(s)))_{s,s' \in S}$. The stationary distribution relative to the exert policy $\pi_E$ is $\rho_E$.

The reward function $R$ is a local measure of the quality of the control. The global quality of the control induced by a policy $\pi$, with respect to a reward $R$, is assessed by the value function $V^\pi_R \in \mathbb{R}^{S}$ which associates to each state the discounted cumulative reward for following policy $\pi$ from this state:
\begin{equation}
V^\pi_R(s) = \E[\sum_{t\geq 0}\gamma^tR(s_t,\pi(s_t))|s_0 = s].
\end{equation}
This long-term criterion is what is being optimized when solving an MDP. Therefore, an optimal policy $\pi^*_R$ is a policy whose value function (the optimal value function $V^*_R$) is greater than that of any other policy, for all states: $\forall \pi, V^*_R\geq V^\pi_R$..

The Bellman evaluation operator $T^\pi_R: \mathbb{R}^{S} \rightarrow  \mathbb{R}^{S}$ is defined by
\begin{equation}
  T^{\pi}_RV = R_\pi + \gamma P_\pi V
  \end{equation}
where $R_\pi$ is $(R(s,\pi(s)))_{s\in S}$. The Bellman optimality operator follows naturally
\begin{equation}
  T^*_RV = \max_\pi T^\pi_RV.
\end{equation}
  Both operators are contractions. The fixed point of the Bellman evaluation operator $T^\pi_R$ is the value function of $\pi$ with respect to reward R:
  \begin{eqnarray}
    V^\pi_R &=& T^\pi_R V^\pi_R\\
    V^\pi_R &=& R_\pi + \gamma P_\pi V^\pi_R.
  \end{eqnarray}
The Bellman optimality operator $T^*_R$ also admits a fixed point, the optimal value function $V_R^*$ with respect to reward $R$.

@@

An optimal policy follows a greedy mechanism with respect to its optimal action-value function $Q^*_R$
\begin{equation}
  \label{eq:greedy}
  \pi^*_R(s)\in\argmax_aQ^*_R(s,a).
\end{equation}

When the state space is too large to allow matricial representation or when the transition probabilites or even the reward function are unknown except through observations gained by interacting with the system, RL or ADP may be used to approximate the optimal control policy~\cite{sutton1998reinforcement}.

We recall that solving the MDP is the direct problem. This contribution aims at solving the inverse one. We observe an expert's deterministic\footnotemark[\value{footnote}] policy $\pi_E$, assuming that there exists some unknown reward $R_E$ for which the expert is optimal. The suboptimality of the expert is an interesting setting that has been discussed for example in \cite{melo2010analysis,syed2010reduction}, but that we are not addressing here. We do not try to find this unknown reward $R_E$ but rather a non trivial reward $R$ for which the expert is at least near-optimal. The trivial reward $0$ is a solution to this ill-posed problem (no reward means that every behavior is optimal). Because of its ill-posed nature, this expression of \emph{Inverse Reinforcement Learning} (IRL) still has to find a satisfactory solution although lots of progress have been made, see Sec.~\ref{sec:related}.
\section{The cascading algorithm}
\label{sec:algo}
Our first step towards a reward function solving the IRL problem is a classification step using a \emph{score function-based multi-class classifier} (SFMC$^2$ for short). This classifier learns a score function $q\in\mathbb{R}^{S\times A}$ that rates the association of a given label $a\in A$ with a certain input $s\in S$. The classification rule $\pi_C\in A^S$ simply selects (one of) the label(s) that achieves the highest score for the given inputs:
\begin{equation}
  \label{eq:greedy2}
\pi_C(s) \in \argmax_a q(s,a).
\end{equation}
For example, \emph{Multi-class Support Vector Machines}~\cite{guermeur2011generic} can be seen as SFMC$^2$ algorithms, the same can be said of the structured margin approach \cite{taskar2005learning} which we consider in the experimental setting. Other algorithms may be envisionned (see Sec.~\ref{subsec:instanciation}).

Given a dataset $D_C = \{(s_i,a_i=\pi^E(s_i))_i\}$ of actions $a_i$ (deterministically) chosen by the expert on states $s_i$, we train such a classifier. The classification policy $\pi_C$ is not the end product we are looking for (that would be mere imitation of the expert, not IRL). What is of particular interest to us is the score function $q$ itself. One can easily notice the similarity between Eq.~\eqref{eq:greedy2} and Eq.~\eqref{eq:greedy} that describes the relation between the optimal policy in an MDP and its optimal action-value function. It is only natural to view the score function $q$ of the classifier as some kind of optimal action-value function for the classifier policy $\pi_C$. By inversing the Bellman equation Eq.~\eqref{eq:bellman1} with $q$ in lieu of $Q^\pi_R$, one gets $R^C$, the reward function relative to our score/action-value function $q$:
\begin{equation}
  \label{eq:rc}
  R^C(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi_C(s')).
\end{equation}
As we wish to solve the general, or approximate, IRL problem where the transition probabilities $P$ are unknown, our reward function $R^C$ will be approximated with the help of information gathered by interacting with the system. We assume that another dataset $D_R = \{(s_j,a_j,s'_j)_j\}$ is available where $s'_j$ is the state an agent taking action $a_j$ in state $s_j$ transitionned to. Action $a_j$ needs not be chosen by any particular policy. The dataset $D_R$ brings us information about the dynamics of the system. From it, we construct datapoints
\begin{equation}
  \label{eq:rj}
  \{\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))\}_j.
\end{equation}
As $s'_j$ is sampled accordingly to $P(\cdot|s_j,a_j)$ the constructed datapoints help building a good approximation of $R^C(s=s_j,a=a_j)$. A regressor (a simple least-square approximator can do but other solutions such as Gaussian Processes \cite{rasmussen2006gaussian} could also be envisionned) is then fed the datapoints $((s_j,a_i),\hat r_j)$ to obtain $\hat R^C$, a generalisation of $\{((s_j,a_j),\hat r_j)_j\}$ over the whole state-action space. The complete algorithm is given in Alg.~\ref{algo:cascading}.
\begin{algorithm}%[H]
    %\small
  \caption{CSI algorithm}
  \label{algo:cascading}
  \emph{\textbf{Given}} a training set $D_C=\{(s_i,a_i=\pi^E(s_i))\}_{1\leq i \leq D}$ and another training set $D_R=\{(s_{j},a_{j},s'_{j})\}_{1\leq j \leq D'}$\;\\
  \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$\;\\
  \emph{\textbf{Learn}} a reward function $\hat R^C$ from the dataset $\{((s_{j},a_{j}),\hat{r}_j)\}_{1\leq j \leq D'}$, $\forall (s_j,a_j,s'_j) \in D_R,\hat{r}_j=q(s_{j},a_{j})-\gamma q(s'_{j},\pi_C(s'_{j}))$\;\\
  \emph{\textbf{Output}} the reward function $\hat R^{C}$ \;
\end{algorithm}

Cascading two supervised approaches like we do is a way to inject the MDP structure in the resolution of the problem. Indeed, mere classification only takes into account information from the expert (i.e., which action goes with which state) whereas using the Bellman equation in the expression of $\hat r_j$ makes use of the information lying in the transitions $(s_j,a_j,s'_j)$, namely information about the transition probabilities $P$. The final regression step is just a way to generalize this information about $P$ to the whole state-action space in order to have a well-behaved reward function.

@@

In this section, we prove that the deterministic expert policy $\pi_E$ is near optimal for the reward $\hat{R}^C$ the regression step outputs.
More formally, we prove that $\E_{s\sim\rho_E}[V^*_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]$ is bounded by a term that depends on:
\begin{itemize}
  \item the classification error
\begin{equation}  
\epsilon_C=\E_{s\sim\rho_E}[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}];
\end{equation}
\item the regression error
  \begin{equation}
    \epsilon_R=\max_{\pi\in A^S}\|\epsilon_\pi^R\|_{1,\rho_E},
  \end{equation}
  with:
  \begin{itemize}
  \item the subscript notation already used in Sec.~\ref{sec:background} meaning that, given an $X\in\mathbb{R}^{S\times A}$, $\pi\in A^S$, and $a\in A$,  $X_\pi\in\mathbb{R}^S$ and $X_a\in\mathbb{R}^S$ are repectively such that: $\forall s\in S, X_\pi(s)=X(s,\pi(s))$ and $\forall s\in S, X_a(s)=X(s,a)$ ;
  \item $\epsilon^R_\pi=R^C_\pi-\hat{R}^C_\pi$ ;
  \item $\|.\|_{1,\mu}$ the $\mu$-weighted $L_1$ norm: $\|f\|_{1,\mu} = \E_{x\sim \mu}[|f(x)|]$ ;
  \end{itemize}
\item the concentration coefficient $C_* = C_{\hat \pi_C}$ with:
  \begin{itemize}
  \item $C_{\pi}=(1-\gamma)\sum_{t\geq0}\gamma^tc_{\pi}(t), \text{ with } c_{\pi}(t)=\max_{s\in S}\frac{(\rho_E^TP^t_\pi)(s)}{\rho_E(s)}$ ;
  \item $\hat \pi_C$, the optimal policy for the reward $\hat R^C$ output by the algorithm ;
  \end{itemize}
  The constant $C_*$ could be estimated \emph{a posteriori} ( after $\hat{R}^C$ is computed). We could upper bound $C_*$ by a more usual and general concentration coefficient but $C_*$ gives a tighter final result (roughly, if $\hat{\pi}_C\approx\pi_E$ then $C_* \approx 1$).
\item $\Delta q$, defined by 
  \begin{eqnarray}
    \Delta q &= &\max_{s\in S}(\max_{a\in A}q(s,a)-\min_{a\in A}q(s,a))\\
    &=&\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a)),
    \end{eqnarray} which could be normalized to $1$ without loss of generality. The range of variation of $q$ induces the one of $R^C$, $\hat{R}^C$ and $V^{\pi_C}_{\hat{R^C}}$.
  \end{itemize}

>>>>

% Created 2012-10-15 lun. 14:49
\documentclass[smallextended]{svjour3}
\smartqed 
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage[integrals]{wasysym}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subfigure}
\tolerance=1000
%\usepackage{amsmath}
%\usepackage{amsthm}
%\newtheorem{definition}{Definition}
%\newtheorem{theorem}{Theorem}
\newtheorem{cor}{Corollary}
\usepackage{dsfont}
\usepackage{mathtools}
\mathtoolsset{showonlyrefs=true}
\providecommand{\alert}[1]{\textbf{#1}}
\newcommand{\E}{\mathbf{E}}
\newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
\newcommand{\Card}{\operatorname*{Card}} %\operatorname* pour les op. pouvant admettre des limites...
\begin{document}
\title{A cascaded supervised learning approach to inverse reinforcement learning}
\author{Edouard Klein$^{1,2}$ \and Bilal Piot $^{2,3}$\and Matthieu Geist $^{2}$\and Olivier Pietquin$^{2,3}$}
\titlerunning{Cascading approach to IRL}
\institute{
 1. ABC Team\\
 LORIA-CNRS, France.\\
\and
2. Supélec\\
 IMS-MaLIS Research group, France\\
 \texttt{firstname.lastname@supelec.fr}\\
\and 
3. UMI 2958 (GeorgiaTech-CNRS)\\
France\\
}
\date{\today}

@@

\begin{abstract}
  This paper considers the Inverse Reinforcement Learning (IRL) problem, that is inferring a reward function for which a demonstrated expert policy is optimal.
We propose to break the IRL problem down into two generic Supervised Learning steps: this is the Cascaded Supervised IRL (CSI) approach. A classification step that defines a score function is followed by a regression step providing a reward function.
A theoretical analysis shows that the demonstrated expert policy is near-optimal for the computed reward function.
Not needing to repeatedly solve a Markov Decision Process (MDP) and the ability to leverage existing techniques for classification and regression are two important advantages of the CSI approach. It is furthermore empirically demonstrated to work with only transitions sampled according to the expert policy, up to the use of some heuristics. This is exemplified on two classical benchmarks (a highway driving simulator and an inverted pendulum simulator).
  \end{abstract}
\section{Introduction}
\label{sec-2}
Sequential decision making consists in choosing the appropriate action given the available data in order to maximize a certain criterion. When framed in a Markov Decision Process (MDP) (see Sec.~\ref{sec:background}), (Approximate) Dynamic programming ((A)DP) or Reinforcement Learning (RL) are often used to solve the problem by maximizing the expected sum of discounted rewards. The Inverse Reinforcement Learning (IRL)~\cite{russell1998learning} problem, which is addressed here, aims at inferring a reward function for which a demonstrated expert policy is optimal.

IRL is one of many ways to perform Apprenticeship Learning (AL): imitating a demonstrated expert policy, without explicitely looking for the reward function. The reward function nevertheless is of interest in its own right. As mentionned in \cite{russell1998learning}, its semantics can be analyzed in biology or econometrics for instance. Practically, the reward can be seen as a succinct description of a task. Discovering it removes the coupling between understanding the task and learning how to fulfill it that exists in AL. IRL allows the use of (A)DP or RL techniques to learn the task from the computed reward function. Although some approaches use it as a byproduct, most AL algorithm makes no use of the reward function, or have nothing to do with IRL altogether. A very straightforward non-IRL way to do AL is for example to use a multi-class classifier to directly learn the expert policy. We provide in the experiments (Sec.~\ref{sec:experiments}) a comparison between AL and IRL algorithms by using IRL as a way to do AL.

A lot of existing approaches in either IRL or IRL-based AL need to repeatedly solve the underlying MDP to find the optimal policies of intermediate reward functions. Thus, their performance depends strongly on the quality of the associated subroutine. Consequently, they suffer from the same challenges of scalability, data scarcity, etc., as RL and (A)DP. In order to avoid \emph{repeatedly} solving such problems, we adopt a different point of view.

Having in mind that there is a one to one relation between a reward function and its associated optimal action-value function (via the Bellman equation, see Eq.~\eqref{eq:bellman1}), it is worth thinking of a method able to output an action-value function for which the greedy policy is the demonstrated expert policy. Thus, the demonstrated expert policy will be optimal for the related reward function. We propose to use a score function-based multi-class classification step (see Sec.~\ref{sec:algo}) to infer a score function. Besides, in order to retrieve via the Bellman equation the reward associated to the score function computed by the classification step, we introduce a regression step (see Sec.~\ref{sec:algo}). That is why the method is called the Cascaded Supervised Inverse reinforcement learning (CSI). This method is analyzed in Sec.~\ref{sec:analysis}, where it is shown that the demonstrated expert policy is near-optimal for the reward the regression step outputs.

This algorithm does not need to iteratively solve an MDP and needs only sampled transitions from expert and non-experts policies as inputs. Moreover, up to the use of some heuristics (see Sec.~\ref{subsubsec:heuristics}), the algorithm is able to be trained only with transitions sampled from the demonstrated expert policy. A specific instantiation of CSI (proposed Sec.~\ref{subsec:instanciation}) is tested on an inverted pendulum simulator (Sec.~\ref{subsec:IP}) and a car driving simulator (Sec.~\ref{subsec:highway}) where we compare it with an AL algorithm \cite{abbeel2004apprenticeship}, with a pure classification algorithm \cite{taskar2005learning} and with a recent successfull IRL method \cite{klein2012scirl} as well as with a random baseline.

@@

We work with finite MDPs \cite{puterman1994markov}, that is tuples $\{S,A,P,R,\gamma\}$. The state space is noted $S$, $A$ is a finite action space, $R\in\mathbb{R}^{S\times A}$ is a reward-function, $\gamma\in ]0,1[$ is a discount factor and $P\in \Delta_{S}^{S\times A}$ is the Markovian dynamics of the MDP. Thus, for each $(s,a)\in S\times A$, $P(.|s,a)$ is a distribution over $S$ and $P(s'|s,a)$ is the probability to reach $s'$ by choosing action $a$ in state $s$. At each time step $t$, the agent uses the information encoded in the state $s_t\in S$ in order to choose an action $a_t \in A$ according to a (deterministic\footnote{We restrict ourselves here to deterministic policies, but the loss of generality is minimal as there exists at least one optimal deterministic policy.}) policy $\pi\in A^S$. The agent then steps to a new state $s_{t+1}\in S$ according to the Markovian transition probabilities $P(s_{t+1}|s_t,a_t)$. The stationary distribution over the states $\rho_\pi$ induced by a policy $\pi$ satisfies
\begin{equation}
  \rho_\pi^TP_\pi = \rho_\pi^T,
  \end{equation}
where $P_\pi$ is $(P(s'|s,\pi(s)))_{s,s' \in S}$. The stationary distribution relative to the exert policy $\pi_E$ is $\rho_E$.

The reward function $R$ is a local measure of the quality of the control. The global quality of the control induced by a policy $\pi$, with respect to a reward $R$, is assessed by the value function $V^\pi_R \in \mathbb{R}^{S}$ which associates to each state the expected discounted cumulative reward for following policy $\pi$ from this state:
\begin{equation}
V^\pi_R(s) = \E[\sum_{t\geq 0}\gamma^tR(s_t,\pi(s_t))|s_0 = s].
\end{equation}
This long-term criterion is what is being optimized when solving an MDP. Therefore, an optimal policy $\pi^*_R$ is a policy whose value function (the optimal value function $V^*_R$) is greater than that of any other policy, for all states: $\forall \pi, V^*_R\geq V^\pi_R$.

The Bellman evaluation operator $T^\pi_R: \mathbb{R}^{S} \rightarrow  \mathbb{R}^{S}$ is defined by
\begin{equation}
  T^{\pi}_RV = R_\pi + \gamma P_\pi V
  \end{equation}
where $R_\pi$ is $(R(s,\pi(s)))_{s\in S}$. The Bellman optimality operator follows naturally:
\begin{equation}
  T^*_RV = \max_\pi T^\pi_RV.
\end{equation}
  Both operators are contractions. The fixed point of the Bellman evaluation operator $T^\pi_R$ is the value function of $\pi$ with respect to reward R:
  \begin{eqnarray}
    V^\pi_R &=& T^\pi_R V^\pi_R \Leftrightarrow V^\pi_R = R_\pi + \gamma P_\pi V^\pi_R.
  \end{eqnarray}
The Bellman optimality operator $T^*_R$ also admits a fixed point, the optimal value function $V_R^*$ with respect to reward $R$.

@@

An optimal policy follows a greedy mechanism with respect to its optimal action-value function $Q^*_R$:
\begin{equation}
  \label{eq:greedy}
  \pi^*_R(s)\in\argmax_aQ^*_R(s,a).
\end{equation}

When the state space is too large to allow matricial representations or when the transition probabilites or even the reward function are unknown except through observations gained by interacting with the system, RL or ADP may be used to approximate the optimal control policy~\cite{sutton1998reinforcement}.

We recall that solving the MDP is the direct problem. This contribution aims at solving the inverse one. We observe trajectories drawn from an expert's deterministic\footnotemark[\value{footnote}] policy $\pi_E$, assuming that there exists some unknown reward $R_E$ for which the expert is optimal. The suboptimality of the expert is an interesting setting that has been discussed for example in \cite{melo2010analysis,syed2010reduction}, but that we are not addressing here. We do not try to find this unknown reward $R_E$ but rather a non trivial reward $R$ for which the expert is at least near-optimal. The trivial reward $0$ is a solution to this ill-posed problem (no reward means that every behavior is optimal). Because of its ill-posed nature, this expression of \emph{Inverse Reinforcement Learning} (IRL) still has to find a satisfactory solution although lots of progress have been made, see Sec.~\ref{sec:related}.
\section{The cascading algorithm}
\label{sec:algo}
Our first step towards a reward function solving the IRL problem is a classification step using a \emph{score function-based multi-class classifier} (SFMC$^2$ for short). This classifier learns a score function $q\in\mathbb{R}^{S\times A}$ that rates the association of a given label $a\in A$ with a certain input $s\in S$. The classification rule $\pi_C\in A^S$ simply selects (one of) the label(s) that achieves the highest score for the given inputs:
\begin{equation}
  \label{eq:greedy2}
\pi_C(s) \in \argmax_a q(s,a).
\end{equation}
For example, \emph{Multi-class Support Vector Machines}~\cite{guermeur2011generic} can be seen as SFMC$^2$ algorithms, the same can be said of the structured margin approach \cite{taskar2005learning} which we consider in the experimental setting. Other algorithms may be envisionned (see Sec.~\ref{subsec:instanciation}).

Given a dataset $D_C = \{(s_i,a_i=\pi_E(s_i))_i\}$ of actions $a_i$ (deterministically) chosen by the expert on states $s_i$, we train such a classifier. The classification policy $\pi_C$ is not the end product we are looking for (that would be mere imitation of the expert, not IRL). What is of particular interest to us is the score function $q$ itself. One can easily notice the similarity between Eq.~\eqref{eq:greedy2} and Eq.~\eqref{eq:greedy} that describes the relation between the optimal policy in an MDP and its optimal action-value function. The score function $q$ of the classifier can thus be viewed as some kind of optimal action-value function for the classifier policy $\pi_C$. By inversing the Bellman equation Eq.~\eqref{eq:bellman1} with $q$ in lieu of $Q^\pi_R$, one gets $R^C$, the reward function relative to our score/action-value function $q$:
\begin{equation}
  \label{eq:rc}
  R^C(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi_C(s')).
\end{equation}
As we wish to solve the general, or approximate, IRL problem where the transition probabilities $P$ are unknown, our reward function $R^C$ will be approximated with the help of information gathered by interacting with the system. We assume that another dataset $D_R = \{(s_j,a_j,s'_j)_j\}$ is available where $s'_j$ is the state an agent taking action $a_j$ in state $s_j$ transitionned to. Action $a_j$ needs not be chosen by any particular policy. The dataset $D_R$ brings us information about the dynamics of the system. From it, we construct datapoints
\begin{equation}
  \label{eq:rj}
  \{\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))\}_j.
\end{equation}
As $s'_j$ is sampled accordingly to $P(\cdot|s_j,a_j)$ the constructed datapoints help building a good approximation of $R^C(s=s_j,a=a_j)$. A regressor (a simple least-square approximator can do but other solutions such as Gaussian Processes \cite{rasmussen2006gaussian} could also be envisionned) is then fed the datapoints $((s_j,a_i),\hat r_j)$ to obtain $\hat R^C$, a generalisation of $\{((s_j,a_j),\hat r_j)_j\}$ over the whole state-action space. The complete algorithm is given in Alg.~\ref{algo:cascading}.
\begin{algorithm}%[H]
    %\small
  \caption{CSI algorithm}
  \label{algo:cascading}
  \emph{\textbf{Given}} a training set $D_C=\{(s_i,a_i=\pi_E(s_i))\}_{1\leq i \leq D}$ and another training set $D_R=\{(s_{j},a_{j},s'_{j})\}_{1\leq j \leq D'}$\;\\
  \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$\;\\
  \emph{\textbf{Learn}} a reward function $\hat R^C$ from the dataset $\{((s_{j},a_{j}),\hat{r}_j)\}_{1\leq j \leq D'}$, $\forall (s_j,a_j,s'_j) \in D_R,\hat{r}_j=q(s_{j},a_{j})-\gamma q(s'_{j},\pi_C(s'_{j}))$\;\\
  \emph{\textbf{Output}} the reward function $\hat R^{C}$ \;
\end{algorithm}

Cascading two supervised approaches like we do is a way to inject the MDP structure in the resolution of the problem. Indeed, mere classification only takes into account information from the expert (i.e., which action goes with which state) whereas using the Bellman equation in the expression of $\hat r_j$ makes use of the information lying in the transitions $(s_j,a_j,s'_j)$, namely information about the transition probabilities $P$. The final regression step is a way to generalize this information about $P$ to the whole state-action space in order to have a well-behaved reward function.

@@

In this section, we prove that the deterministic expert policy $\pi_E$ is near optimal for the reward $\hat{R}^C$ the regression step outputs.
More formally, recalling from Sec.~\ref{sec:background} that $\rho_E$ is the stationary distribution of the expert policy, we prove that $\E_{s\sim\rho_E}[V^*_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]$ is bounded by a term that depends on:
\begin{itemize}
  \item the classification error defined as
\begin{equation}  
\epsilon_C=\E_{s\sim\rho_E}[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}];
\end{equation}
\item the regression error defined as 
  \begin{equation}
    \epsilon_R=\max_{\pi\in A^S}\|\epsilon_\pi^R\|_{1,\rho_E},
  \end{equation}
  with:
  \begin{itemize}
  \item the subscript notation already used for $R_\pi$ and $P_\pi$ in Sec.~\ref{sec:background} meaning that, given an $X\in\mathbb{R}^{S\times A}$, $\pi\in A^S$, and $a\in A$,  $X_\pi\in\mathbb{R}^S$ and $X_a\in\mathbb{R}^S$ are repectively such that: $\forall s\in S, X_\pi(s)=X(s,\pi(s))$ and $\forall s\in S, X_a(s)=X(s,a)$ ;
  \item $\epsilon^R_\pi=R^C_\pi-\hat{R}^C_\pi$ ;
  \item $\|.\|_{1,\mu}$ the $\mu$-weighted $L_1$ norm: $\|f\|_{1,\mu} = \E_{x\sim \mu}[|f(x)|]$ ;
  \end{itemize}
\item the concentration coefficient $C_* = C_{\hat \pi_C}$ with:
  \begin{itemize}
  \item $C_{\pi}=(1-\gamma)\sum_{t\geq0}\gamma^tc_{\pi}(t), \text{ with } c_{\pi}(t)=\max_{s\in S}\frac{(\rho_E^TP^t_\pi)(s)}{\rho_E(s)}$ ;
  \item $\hat \pi_C$, the optimal policy for the reward $\hat R^C$ output by the algorithm ;
  \end{itemize}
  The constant $C_*$ could be estimated \emph{a posteriori} (after $\hat{R}^C$ is computed). We could upper bound $C_*$ by a more usual and general concentration coefficient but $C_*$ gives a tighter final result (roughly, if $\hat{\pi}_C\approx\pi_E$ then $C_* \approx 1$).
\item $\Delta q$, defined by 
  \begin{eqnarray}
    \Delta q &= &\max_{s\in S}(\max_{a\in A}q(s,a)-\min_{a\in A}q(s,a))=\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a)),
    \end{eqnarray} which could be normalized to $1$ without loss of generality. The range of variation of $q$ induces the one of $R^C$, $\hat{R}^C$ and $V^{\pi_C}_{\hat{R^C}}$.
  \end{itemize}

\begin{theorem}
\label{thm}
Let $\pi_E$ be the deterministic expert policy, $\rho_E$ its stationary distribution and $\hat{R}^C$ the reward the cascading algorithm outputs. We have:
\begin{equation}
0\leq\E_{s\sim\rho_E}[V^*_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]\leq \frac{1}{1-\gamma}\left(\epsilon_C\Delta q +\epsilon_R(1+C_*)\right).
\end{equation}
\end{theorem}
\begin{proof}
First let's introduce some notations, $q\in\mathbb{R}^{S\times A}$ is the score function output by the classification step, $\pi_C$ is a deterministic classifier policy so that $\forall s \in S,\pi_C(s)\in\argmax_{a\in A}q(s,a)$. We recall the definition of the reward $R^C\in\mathbb{R}^{S\times A}$:
\begin{equation}
\forall (s,a)\in S\times A, R^C(s,a)=q(s,a) -\gamma\sum_{s'\in S}P(s'|s,a)q(s',\pi_C(s')),
\end{equation}
and the definition of $\hat{R}^C\in\mathbb{R}^{S\times A}$ which is the reward function output by the regression step.
